---
title: "Chapter 2 Notes"
format: html
editor: visual
---

## Aaron Freeman \| Feb 1st, 2024

### swcarpentry r-novice tutorial

#### https://swcarpentry.github.io/r-novice-gapminder/04-data-structures-part1.html

### Data Structures

#### Overview

One of R's most powerful features is the ability to deal with tabular data. Such as a csv file. Let's start by making a fake dataset in the data/directory, called feline-data.csv

```r
cats <- data.frame(coat = c("calico", "black", "tabby"),
                   weight = c(2.1, 5.0, 3.2),
                   likes_string = c(1, 0, 1))
```

Now we have to save the cats as a csv. We are setting row.names = FALSE. Use ?write.csv to pull up the help file

```r
write.csv(x = cats, file = "data/feline-data.csv", row.names = FALSE)
```

New contents of the file, feline-data.csv:

```r
coat,weight,likes_string
calico,2.1,1
black,5.0,0
tabby,3.2,1
```

We can load this into R via the following:

```r
cats <- read.csv(file = "Data/feline-data.csv")
cats
```

We can begin exploring our dataset right away, pulling out columns by specifying them using the $ operator:

```r
cats$weight
```

```r
cats$coat
```

We can do other operations

```r
cats$weight + 2
```

```r
paste("my cat is", cats$coat)
```

### Datatypes

Type this to check the data type

```r
typeof(cats$weight)
```

There are 5 main types: double, integer, complex, logical, and character. For historic reasons, double is also called numeric.

```{r}
typeof(3.14)
```

```{r}
typeof(1L) # The L suffix forces the number to be an integer, by default R uses float numbers
```

```{r}
typeof(1+1i)
```

```{r}
typeof(TRUE)
```

```{r}
typeof('banana')
```

